Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-11-04 fstats_pca_lewontin.tex
better discussion

Previous interpretation of PCA in the context of population genetic models have focused on explicit models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{francois2021}. My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs. One consequence is that the results here are not impacted by sample ascertainment, sample sizes or number of principal components analyzed, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.

@@

@@

As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}. However, in many regions, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}. This provides a challenge for interpretation; as many $F_3$ and $F_4$ statistics may indicate gene flow. In my example (Figure \ref{fig:f3}A), most Southern European populations are ``admixed'' between Basques and Turkish, but a more accurate model might be one of continuous variation where Basque and Turkish lie on one of multiple gradients; which is more directly visualized with PCA. There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. One example of this is the world foci data set used here, which contains only 33 populations from across the world, and which is well-approximated by two PCs. However, this ascertainment misses a large amount of variation; a more dense sampling would show that in many places human genetic diversity is very gradual and multi-layered \citep{lazaridis2014, peter2020a}. The PCA-based  interpretation offers an alternative that trades interpretability for robustness. Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow. Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.

>>>>

@@

\subsection{Previous Interpretations}
Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2021}. My interpretation here is different in that the utility of PCA is to simplify the geometry of the data. One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.

The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs. Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs.

@@

As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}. However, in many parts of the world, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}. This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness. A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:f3}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians. In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.

This trade-off between $F$-statistics, which allow for significance teste tension between

There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. One example of this is the world foci data set used here, which contains only 33 populations from across the world, and which is well-approximated by two PCs. However, this ascertainment misses a large amount of variation; a more dense sampling would show that in many places human genetic diversity is very gradual and multi-layered \citep{lazaridis2014, peter2020a}. The PCA-based  interpretation offers an alternative that trades interpretability for robustness. Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow. Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.
